---
title: "Predictive Modeling"
author: "Emma Peng"
date: "2025-04-26"
output: 
  html_document:
    toc: true
    toc_float: true 
    toc_depth: 3
---

# <u> **Election Outcomes Modeling** </u>

# **Logistic Regression**

## Step 1: Get the Data

Import filtered dataset with NULL values in key columns already removed, includes 13,346 of the 27,519 rows used for Ballotpedia's analyses & reports. 

```{r}
ballot <- read.csv("ballotpedia_filtered.csv", stringsAsFactors = TRUE)
# str(ballot)
# summary(ballot)

```


## Step 2: Clean the Data

Remove unnecessary columns and factorize columns that may be used as predictors; don't factorize everything given that many of the qualitative variables have dozens of levels if factorized. 

```{r}
library(dplyr)
library(caret)

colnames(ballot)[colnames(ballot) == "X"] <- "stage"


# Get rid of unnecessary columns
ballot <- ballot %>%
  select(-ends_with("_id"))

ballot$id <- NULL
ballot$candidate <- NULL
ballot$first_name <- NULL
ballot$last_name <- NULL
ballot$row_index <- NULL
ballot$votes_against <- NULL
ballot$delegates_pledged <- NULL
ballot$parent_district_name <- NULL
ballot$election_date <- NULL
ballot$election_year <- NULL
ballot$stage_party <- NULL

# Factorize variables that may be used as predictors; use factors with small num of levels
ballot$is_incumbent <- as.factor(ballot$is_incumbent)
ballot$gender <- as.factor(ballot$gender)
ballot$stage <- as.factor(ballot$stage)
ballot$office_branch <- as.factor(ballot$office_branch)

# Create new datafrae with selected predictors
new_ballot <- ballot %>%
  select(total_disbursements, total_contributions, spending_rank, gender, stage, office_branch, is_incumbent, candidate_status)

# Make response variable binary 
new_ballot$candidate_status <- ifelse(new_ballot$candidate_status == "Won", 1, 0) 

# One-hot encode numeric select predictors
factors <- dummyVars(~ . -1, data = new_ballot)
new_ballot <- predict(factors, newdata = new_ballot)
new_ballot<- as.data.frame(new_ballot) 


```

```{r}
predictors <- new_ballot[, !names(new_ballot) %in% "candidate_status"]
target <- new_ballot$candidate_status

# Function to normalize
minmax <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

# Normalize only numeric columns
scaled_predictors <- predictors %>%
  mutate(across(where(is.numeric), minmax))

final_ballot <- cbind(scaled_predictors, candidate_status = target)

```

## Step 3: Split Data

```{r}

set.seed(12345)
train_rows <- sample(1:nrow(final_ballot), .5*nrow(final_ballot))

train_ballot <- final_ballot[train_rows, ]
test_ballot <- final_ballot[-train_rows, ]  

```

## Step 4: Build a Model

Build a simple model to start using only numeric values. 

```{r}

m1 <-glm(candidate_status ~ total_disbursements + total_contributions + spending_rank, data = train_ballot, family = "binomial" )
summary(m1)


```

## Step 5: Predict the Test Data


```{r}
ballot_pred <- predict(m1, test_ballot, type = "response")
summary(ballot_pred)

# Start with threshold of 0.5
predict_binary <- ifelse(ballot_pred < 0.5, 0, 1)
summary(predict_binary)
table(predict_binary)

```

## Step 6: Evaluate the Model

```{r}
confusionMatrix(factor(predict_binary), factor(test_ballot$candidate_status), positive = "1") 
```
## Step 7: Improve the Model

Include factorized categorical predictors

```{r}
m2 <-glm(candidate_status ~ ., data = train_ballot, family = "binomial" )
ballot_pred2 <- predict(m2, test_ballot, type = "response")
summary(ballot_pred2)

# Start with threshold of 0.5
predict_binary2 <- ifelse(ballot_pred2 < 0.5, 0, 1)
summary(predict_binary2)
table(predict_binary2)

confusionMatrix(factor(predict_binary2), factor(test_ballot$candidate_status), positive = "1") 


```
# **KNN**

Create initial model of just numeric predictors. 

```{r}
library(class)

predictors <- c("total_disbursements", "total_contributions", "spending_rank")


pred_knn <- knn(train = train_ballot[, predictors],
                 test = test_ballot[, predictors],     
                 cl = train_ballot$candidate_status,      
                 k = 19)  

confusionMatrix(as.factor(pred_knn), as.factor(test_ballot$candidate_status), positive = "1")   

```


```{r}

pred_knn2 <- knn(
  train = train_ballot[, !names(train_ballot) %in% "candidate_status"], 
  test = test_ballot[, !names(test_ballot) %in% "candidate_status"],
  cl = train_ballot$candidate_status, 
  k = 19
)

confusionMatrix(as.factor(pred_knn2), as.factor(test_ballot$candidate_status), positive = "1")
```

# **ANN**

Again, we can use the same training and testing datasets since ANN also only takes in numeric values.

```{r}
names(train_ballot) <- make.names(names(train_ballot))
```


```{r}
library(neuralnet)
set.seed(12345) 
# m1_ann <- neuralnet(candidate_status ~ ., hidden = 1, threshold = 0.01, stepmax = 1e8, lifesign = "full", data = train_ballot)
```

```{r}
# saveRDS(m1_ann, "ANNModelM1.rds")
m1_ann <- readRDS("ANNModelM1.rds")
```

```{r}
names(test_ballot) <- make.names(names(test_ballot))

pred_ann <- predict(m1_ann, test_ballot)
summary(pred_ann)

binary_pred_ann <- ifelse(pred_ann < 0.3, 0, 1)

confusionMatrix(as.factor(binary_pred_ann), as.factor(test_ballot$candidate_status), positive = "1")
```

# **SVM**

Use vanilladot just for a linear model to start. 

```{r}
library(kernlab)
m1_svm <- ksvm(as.factor(candidate_status) ~ ., data = train_ballot, kernel = "vanilladot")

```

```{r}
pred_svm <- predict(m1_svm, test_ballot) 
confusionMatrix(as.factor(pred_svm), as.factor(test_ballot$candidate_status), positive = "1")

```
Use rbfdot to see if this fits better

```{r}
# m2_svm_prob <- ksvm(as.factor(candidate_status) ~ ., data = test_ballot, kernel = "rbfdot", prob.model = TRUE)
# saveRDS(m2_svm_prob, "svm_probs.rds")
svm_pred_probs <- readRDS("svm_probs.rds") 
svm_probs <- predict(svm_pred_probs, test_ballot, type = "prob")
svm_probs <- svm_probs[, 2] 
svm_binary_pred <- ifelse(svm_probs < 0.5, 0, 1) 
confusionMatrix(as.factor(svm_binary_pred), as.factor(test_ballot$candidate_status), positive = "1")

```
# **Decision Tree**

Create a decision tree for the last predictions input for the stacked model. 

```{r}
library(C50)
tree1 <- C5.0(as.factor(candidate_status) ~ ., data = train_ballot)
plot(tree1)
pred_tree <- predict(tree1, test_ballot, type = "prob")
tree_binary_pred <- ifelse(pred_tree[,2] < 0.5, 0, 1)
confusionMatrix(as.factor(tree_binary_pred), as.factor(test_ballot$candidate_status), positive = "1")

```

# **Stacked Model**

Use the prediction vectors of all the models made to create the second-level stacked model.

```{r}
# Combine the vectors into columns of a dataframe
stack_data <- data.frame(ballot_pred, pred_knn2, pred_ann, svm_probs, pred_tree, test_ballot$candidate_status)

# Check for the right number of rows 
dim(stack_data)
dim(ballot)

# Remove X0 because we only want probabilities for class 1
stack_data$X0 <- NULL

```
```{r}
set.seed(12345)
test_rows_stacked <- sample(1:nrow(stack_data), 0.3*nrow(stack_data))

stacked_test <- stack_data[test_rows_stacked, ]
stacked_train <- stack_data[-test_rows_stacked, ]
```

```{r}
stack_tree <- C5.0(as.factor(test_ballot.candidate_status) ~ ., data = stacked_train)
plot(stack_tree)

```
```{r} 
stack_pred <- predict(stack_tree, stacked_test, type = "prob")

stack_binary_pred <- ifelse(stack_pred[,2] < 0.5, 0, 1)

confusionMatrix(as.factor(stack_binary_pred), as.factor(stacked_test$test_ballot.candidate_status), positive = "1")
```
Summary of the models

```{r}
data.frame(
  Model = c('Logistic', 'KNN', 'ANN', 'SVM', 'Decision Tree', 'Stacked Model'),
  Accuracy = c(0.8723, 0.896, 0.8483, 0.8698, 0.8969, 0.8991),
  Kappa = c(0.739, 0.7904, 0.7001, 0.7334, 0.7914, 0.7971),
  Sensitivity = c(0.7570, 0.8798, 0.9560, 0.7459, 0.8524, 0.8904),
  Specificity = c(0.9706, 0.9098, 0.7566, 0.9753, 0.9348, 0.9065)
)

```

















